---
title: "Lending Club Loan Analysis"
author: "Sahana Ramakrishnan"
date: "2024-07-07"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(sf)
library(usmap)
library(viridis)
```

## --------------------------- MERGING AND CLEANING --------------------------------

```{r}
# 1. Reading CSV files
data2012 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2012.csv")
data2013 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2013.csv")
data2014 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2014.csv")
data2015 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2015.csv")
data2016 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2016.csv")
data2017 <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/data2017.csv")

#2. Combining all data frames into one
combined_df <- bind_rows(data2012, data2013, data2014, data2015, data2016, data2017)

#3. Reading States.csv and states_regions.csv files
states_data <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/States.csv")
states_regions <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/states_regions.csv")

#4. Renaming columns to match for merging
states_data <- states_data |>
  rename(state_name = Geography)

states_regions <- states_regions |>
  rename(state_name = State, state_code = State.Code)

#5. Merging states_data with states_regions to get state code and region info
Stregion <- merge(states_data, states_regions, by = "state_name")

# 6. Merging the combined loan data with the states-region data using the appropriate keys
# to ensure column names match for the merge
final_data <- merge(combined_df, Stregion, by.x = "addr_state", by.y = "state_code")

# 7. Save the final merged data to a CSV file for verification
write.csv(final_data, "D:/NewEng/Week 15/R/Final Case Analysis/full_clean_data.csv", row.names = FALSE)

```

## ------------------------------------ ANALYSIS -----------------------------------------

### 1. Distribution of number of loans by state, regions, and divisions

```{r}
final_data <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/full_clean_data.csv")

#No. of loans by state
loans_by_state <- final_data |>
  group_by(addr_state) |>
  summarise(num_loans = n()) |>
  arrange(desc(num_loans))


# No. of loans by region and division
loans_by_region <- final_data |>
  group_by(Region) |>
  summarise(num_loans = n()) |>
  arrange(desc(num_loans))

loans_by_division <- final_data |>
  group_by(Division) |>
  summarise(num_loans = n()) |>
  arrange(desc(num_loans))


#No. of loans per capita
loans_per_capita <- final_data |>
  group_by(state_name) |>
  summarise(num_loans = n(), population = mean(population)) |>
  mutate(loans_per_capita = (num_loans / population)*100000)


# Checking for missing states
missing_states <- anti_join(states_regions, final_data, by = c('state_code' = 'addr_state'))


  
#Printing results:
print(loans_by_state) 
print(loans_by_region) 
print(loans_by_division) 
print(loans_per_capita)
print(missing_states) 
```

-   California, New York, and Texas have issued the highest no. of loans. This is expected due to their higher population.
-   South, West, and Northeast are the most issued loan regions.
-   Pacific, and South Atlantic are the most issued loan divisions
-   States with smaller population but relatively high number of loans show higher loans per capita.
-   Eg: District of Columbia followed by Connecticut, have high loans per capita values.
-   Conversely, even though California has the highest absolute number of loans, its loans per capita is moderate compared to its population size.
-   Maine and North Dakota are the missing states which could be due to regulatory reasons, as certain states have stricter regulations that may limit or prevent lending activity

### 2. Average loan amount by states and divisions

```{r}
# average loan amount by state
avg_loan_by_state <- final_data |>
  group_by(addr_state) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE)) |>
  arrange(avg_loan_amnt)

# average loan amount by division
avg_loan_by_division <- final_data |>
  group_by(Division) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE)) |>
  arrange(avg_loan_amnt)

# Print results
print(avg_loan_by_state)
print(avg_loan_by_division)
```

-   Mississippi has the highest average loan amount by state
-   Idaho has the highest average loan amount by state
-   New England has the highest average loan amount by division
-   West North Central has the lowest average loan amount by division

### 3. Average Interest Rate and Loan Amount by Grade

```{r}
avg_interest_loan_by_grade <- final_data |>
  group_by(grade) |>
  summarise(avg_int_rate = mean(as.numeric(gsub("%", "", int_rate)), na.rm = TRUE),
            avg_loan_amnt = mean(loan_amnt, na.rm = TRUE))

print(avg_interest_loan_by_grade)
```

-   Grade A loans have the lowest average interest rate at 7.74%.

-   Grade G loans have the highest average interest rate at 25.5%.

-   There is a clear pattern where lower loan grades (from A to G) are associated with higher avg. interest rates.

-   This is consistent with the risk-based pricing model, where loans with higher risk (lower grades) are charged higher interest rates to compensate for the increased risk of default.

-   The average loan amount generally increases as the loan grade decreases.

-   The trend indicates that borrowers with lower grades tend to take out larger loan amounts.

-   This could be because borrowers with lower creditworthiness might need larger sums, possibly due to fewer available credit options elsewhere.

### 4. Frequency Distribution by Year

```{r}
freq_dist_by_year <- final_data |>
  group_by(addr_state, issue_d) |>
  summarise(num_loans = n(),
            avg_loan_amnt = mean(loan_amnt, na.rm = TRUE),
            avg_int_rate = mean(as.numeric(gsub("%", "", int_rate)), na.rm = TRUE))

print(freq_dist_by_year)
```

- Early years (2012-2013) show significant fluctuations in loan numbers, followed by a steady increase from 2013 onwards, reflecting Lending Club's growing popularity and lending capacity.

- Noticeable peaks around the end of each year suggest increased borrowing during holiday season

- The increasing trend post-2013, the impact of the IPO, indicating increased capital and market confidence, with a general upward trend from 2015 onwards.

### 5. Relationship between population size and median income

```{r}
pop_loan_relationship <- final_data |>
  group_by(state_name) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE),
            population = mean(population))

correlation_pop_loan <- cor(pop_loan_relationship$avg_loan_amnt, pop_loan_relationship$population, use = "complete.obs")

print(pop_loan_relationship)
print(correlation_pop_loan)
```

- Correlation Analysis: The corr. coeff between population size of a state and avg. loan amount given is 0.123, indicating a very weak positive relationship between the two variables.

- i.e. there is a slight tendency for states with larger populations to have slightly higher average loan amounts, but this relationship is not strong.

- State-Level Observations: States with large populations, like California and Florida, do not exhibit drastically higher average loan amounts compared to states with smaller populations

- This further supports the observation that the relationship between population size and average loan amount is weak.

```{r}
# Relationship between loan grade and median income level
income_grade_relationship <- final_data |>
  group_by(state_name, grade) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE),
            median_income = mean(Median_income_Households, na.rm = TRUE))

correlation_income_grade <- income_grade_relationship |>
  group_by(grade) |>
  summarise(correlation = cor(avg_loan_amnt, median_income, use = "complete.obs"))


print(income_grade_relationship)
print(correlation_income_grade)
```

- The overall trend shows positive corr. for most loan grades, suggesting that higher median income levels are associated with higher average loan amounts.
- Grades with higher positive correlations (A, B, and D) may indicate that borrowers with better creditworthiness (higher grades) in wealthier states tend to secure larger loans.
- The negative or low corr. in lower grades (F and G) might suggest that these loans are either less influenced by median income or that these grades are more prevalent in states with varying economic conditions.

### 6. Open End Analysis

```{r}
# Trends in loan amounts over time
loan_trends <- final_data |>
  group_by(issue_d) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE))

print(loan_trends)
```

-   Based on the above loan_trends data, the avg. loan amount issued by Lending Club varied significantly over the months in 2013. The highest avg. loan amount was issued in Feb 2013 at $15,894, while the lowest was in Dec 2012 at $14,161.
-   This fluctuation suggests that the demand for loans or the lending criteria may have changed throughout the year.

## ------------------------ VISUALIZATION ----------------------

### 1. Plot for Interest Rates and Grade/Loan 

```{r}
# Ensuring the interest rate is numeric
final_data <- final_data |>
  mutate(int_rate_num = as.numeric(gsub("%", "", int_rate)))

# box plot of interest rates by loan grade
ggplot(final_data, aes(x = grade, y = int_rate_num)) +
  geom_boxplot() +
  labs(title = "Interest Rates by Loan Grade", x = "Loan Grade", y = "Interest Rate (%)") +
  theme_minimal()

# Printing the plot
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/interest_rates_by_grade.png")

# Describing the pattern
summary_by_grade <- final_data |>
  group_by(grade) |>
  summarise(avg_int_rate = mean(int_rate_num, na.rm = TRUE),
            median_int_rate = median(int_rate_num, na.rm = TRUE),
            min_int_rate = min(int_rate_num, na.rm = TRUE),
            max_int_rate = max(int_rate_num, na.rm = TRUE))

print(summary_by_grade)
```

- higher loan grades (such as A and B) tend to have lower interest rates, while lower loan grades (such as E, F, and G) have higher interest rates.
- This pattern is consistent with the risk-based pricing model, where borrowers with higher grades receive loans at lower interest rates, while those with lower grades are charged higher rates to compensate for the higher risk of default.

### 2. Map of US States and color code the map with avg amount of loans given

```{r}
# average loan amount by state
avg_loan_by_state <- final_data |>
  group_by(addr_state) |>
  summarise(avg_loan_amnt = mean(loan_amnt, na.rm = TRUE))

# Ensuring state abbreviations are uppercase for consistency
avg_loan_by_state <- avg_loan_by_state |>
  rename(state = addr_state)


# Creating map
plot_usmap(data = avg_loan_by_state, values = "avg_loan_amnt", regions = "states") +
  scale_fill_viridis(option = "C", na.value = "white", name = "Avg Loan Amount ($)") +
  theme_void() +
  labs(title = "Average Loan Amount by State")

# saving the plot
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/average_loan_amount_by_state.png")
```

- using plot_usmap function to plot the data with average loan amounts, applying the viridis color scale for visual clarity

### 3. Relationship between annual income and loan amount obtained from lending club

```{r, out.height=600, out.width=800}
# Ensuring annual income and loan amount are numeric
final_data <- final_data |>
  mutate(annual_inc = as.numeric(annual_inc),
         loan_amnt = as.numeric(loan_amnt))

# scatter plot
ggplot(final_data, aes(x = annual_inc, y = loan_amnt)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Annual Income and Loan Amount",
       x = "Annual Income ($)",
       y = "Loan Amount ($)") +
  theme_minimal() 
# Saving the plot
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/income_vs_loan_amount.png")
```

- Using ggplot2 to create scatter plot with annual_inc on x-axis and loan_amnt on y-axis.
- Adding points with some transparency (alpha = 0.5) and a linear regression line geom_smooth(method = "lm") to show the trend.

### 4. Relationship between the length of employment and amount of loan obtained 

```{r, out.height=600, out.width=800}
# Cleaning employment length data
final_data <- final_data |>
  mutate(emp_length = case_when(
    emp_length == "< 1 year" ~ "0 years",
    emp_length == "10+ years" ~ "10 years",
    TRUE ~ emp_length
  )) |>
  mutate(emp_length = as.numeric(gsub(" years", "", emp_length)))

# box plot
ggplot(final_data, aes(x = as.factor(emp_length), y = loan_amnt)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(title = "Length of Employment and Loan Amount",
       x = "Length of Employment (years)",
       y = "Loan Amount ($)") +
  theme_minimal() 
#saving
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/emp_length_vs_loan_amount.png")
```

- cleaning emp_length data to make it numeric. 
- Converting "< 1 year" to "0 years" and "10+ years" to "10 years".
- removing the " years" text and convert the column to numeric.
- Using ggplot2 to create box plot with emp_length on x-axis and loan_amnt on y-axis.
- Using as.factor to ensure emp_length is treated as a categorical variable.

### 5. Regional map for any interesting relationship

```{r}
# interest rate to be numeric
final_data <- final_data |>
  mutate(int_rate_num = as.numeric(gsub("%", "", int_rate)))


# average interest rate and median income by region
region_data <- final_data |>
  group_by(Region) |>
  summarise(avg_int_rate = mean(int_rate_num, na.rm = TRUE),
            median_income = mean(Median_income_Households, na.rm = TRUE)) |>
  ungroup()

# Spatial data for US states and add region information
states_regions <- read.csv("D:/NewEng/Week 15/R/Final Case Analysis/states_regions.csv")
states_regions <- states_regions |>
  rename(state_code = State.Code)


# Merging region data with states map
states_regions <- states_regions |>
  left_join(region_data, by = "Region")

# Renaming state to state_code for plot_usmap compatibility
states_regions <- states_regions |>
  rename(state = State)

# Map
plot_usmap(data = states_regions, values = "avg_int_rate", regions = "states") +
  scale_fill_viridis(option = "C", na.value = "white", name = "Avg Interest Rate (%)") +
  theme_void() +
  labs(title = "Average Interest Rate by Region with Median Income") +
  theme(legend.position = "right")

# Saving
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/regional_map_interest_rate_income.png")

# Creating a scatter plot to visualize the relationship
ggplot(region_data, aes(x = median_income, y = avg_int_rate)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Median Income vs Avg Interest Rate (region)",
       x = "Median Income ($)",
       y = "Average Interest Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(t = 20, b = 10)),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20))

# Save the scatter plot
ggsave("D:/NewEng/Week 15/R/Final Case Analysis/income_vs_interest_rate.png")
```
- The scatter plot reveals a slight negative trend between median income and average interest rate, indicating that regions with higher median incomes tend to have slightly lower interest rates. 
- The regional map highlights significant geographic differences, with higher interest rates in the Northeast and West, and lower rates in the South and Midwest. 
- These insights suggest that Lending Club may adjust interest rates based on regional economic conditions. 
- These patterns can help in tailoring lending strategies and improving regional targeting.
























